A statistical comparison of influenza intensity thresholds obtained from the moving epidemic and WHO methods

### 2021-02-10 Draft/Draft_MEM.tex
Updating drafts.

\newcommand{\sd}{s}
\newcommand{\mean}{\bar{Y}}

@@

\title{A bootstrap-based comparison of ILI intensity thresholds from the moving epidemic and WHO methods}
\author{Johannes Bracher, Jonas Littek}
\date{ \small
Karlsruhe Institute of Technology (KIT), Chair of Statistics and Econometrics}

@@

\abstract{The moving epidemic method (MEM) and the WHO method are widely used approaches to determine intensity levels for seasonal influenza and influenza-like illness (ILI). They are conceptually similar, but differ in two aspects. Firstly, the MEM involves a log-transformation of incidence data, while the WHO method typically operates on the original scale of the data. Secondly, the MEM method by default uses more than one observation from each past season, with a fixed total number of observations to include. The WHO method, on the other side, uses only the single highest value from each season to determine peak intensity thresholds. We provide statistical arguments and a simulation study to assess the impact of these choices on the resulting intensity thresholds and empirical exceedance proportions. The simulation study is based on a bootstrap approach using historical ILI data from France, Spain, Switzerland and the United States. We find that using data on the original scale leads to a too large proportion of seasons being classified as high or very high intensity intensity, which can to some degree be mitigated by using a logarithmic transformation. When fixing the total number of observations to include as in the MEM, intensity thresholds tend to increase as more historical data becomes available. When data is only available for few seasons, there is a rather large chance of classifying a new season as high or very high intenstity. We therefore advocate always using one observation per season with a log transformation, i.e.\ a hybrid between the default settings of the MEM and WHO methods. This adaptation is possible within an existing software implementation of the MEM approach and thus easy to apply.}

@@

Following the 2009 influenza H1N1 pandemic, the need for a rapid assessment tool for influenza intensity was recognized. In a report, the Review Committee on the Functioning of the
International Health Regulations and on Pandemic Influenza (H1N1) recommended that member states apply and evaluate their methods for severity assessment every year, thus improving their pandemic preparedness \citep[p.118]{WHO2011}. In the subsequent WHO Pandemic Infuenza Severity Assessment (PISA) guideline (\citeyear{WHO2017}), two specific statistical methods are recommended to determine influenza intensity thresholds. Firstly, the so-called \textit{WHO method}, already described in a previous document \citep{WHO2014} and in more detail in the PISA guide; and secondly, the \textit{moving epidemic method} (MEM; \citealt{Vega2013, Vega2015}) which is also recommended by the European Centre for Disease Prevention and Control (e.g., \citealt{ECDC2017}).

In particular the MEM approach has in the meantime been adopted by numerous national public health agencies (see references in Section \ref{sec:recent_applications}) and seems to deliver practically useful thresholds. Its application has been facilitated by the availablility of a user-friendly and well-documented \texttt{R} package (\texttt{mem}; \citealt{Lozano2020}), which can even be used via a graphical user interface \citep{Lozano2018}. The statistical properties of the MEM and WHO methods, however, have not yet been studied in detail. We obtain some analytical results, complemented by simulation experiments. In order to ensure a close link to real influenza patterns, the latter are based on bootstrap re-sampling of historical surveillance data. This way, we provide an improved understanding and intuition for the consequences of different implementation choices.

The article is structured as follows. Section \ref{sec:definitions} provides concise definitions of the MEM and WHO methods, highlighting two important differences between these conceptually similar approaches. Section \ref{sec:recent_applications} consists of an overview of published applications of the MEM and WHO methods. This is followed by a brief examination of statistical properties of the two methods in Section \ref{sec:analytical}. In Section \ref{sec:simulation} we describe our simulation approach and the results based on historical data from France, Spain, Switzerland and the United States. Section \ref{sec:discussion} concludes with a discussion.

@@

In the following we focus on the construction of thresholds  for peak intensity, framing the MEM and WHO methods as different special cases of the same general approach. Details on the construction of baseline epidemic thresholds from pre-season observations are omitted. We assume that thresholds are based on data from $m$ past seasons and applied to the ($m + 1$)-th season. \cite{Vega2015} recommend to use $5 \leq m \leq 10$ seasons for the MEM to ensure a recent data basis. Thresholds for for a given intensity measure $X$ are then obtained via the following steps:

\begin{enumerate}
\item Within each historical season $j = 1, \dots, m$ order all observations in decreasing order, denoting the $i$-th largest observation from season $j$ by $X^{(j)}_i$.
\item Select the $n$ largest observations from each of the $m$ past seasons to construct a reference set $\mathcal{X} = \{X_i^{(j)}: j = 1, \dots, m; i = 1, \dots, n\}$.
\item Apply an (invertible) transformation $Y_i^{(j)} = f(X_i^{(j)})$ to all members of the reference set $\mathcal{X}$ to obtain a reference set $\mathcal{Y}$ of transformed historical observations.
\item Assume that the transformed values in $\mathcal{Y}$ come from a normal distribution and compute estimates $\mean, \sd$ of its mean and standard deviation.
\item Define intensity thresholds on the transformed scale as quantiles of the normal distribution N$(\mean, \sd^2)$, i.e.\ compute
\begin{equation}
q_{Y, \alpha} = \mean + z_\alpha \sd, \label{eq:def_q}
\end{equation}
where $z_\alpha$ is the $\alpha$ quantile of the standard normal distribution. A common choice for both methods \citep{WHO2017}, also implemented as the default in the \texttt{R} package \texttt{mem} \citep{Lozano2020} is to use
\begin{itemize}
\item the 40th percentile $q_{Y, 0.4} = \mean - 0.25 \sd$ as the threshold between low and medium intensity;
\item the 90th percentile $q_{Y, 0.9} = \mean + 1.28 \sd$ as the threshold between medium and high intensity;
\item the 97.5th percentile $q_{Y, 0.975} = \mean + 1.96\sd$ as the threshold between high and very high intensity.
\end{itemize}
\item Obtain thresholds on the original scale by applying the inverse transformation, i.e. setting $q_{0.4} = f^{-1}(q_{Y, 0.4})$, $q_{0.9} = f^{-1}(q_{Y, 0.9})$, $q_{0.975} = f^{-1}(q_{Y, 0.975})$.
\end{enumerate}

\noindent The MEM and WHO methods are special cases of this procedure with the following specifications:
\begin{itemize}
\item In the WHO method, the standard procedure is to set $f$ to the identity function (i.e.\ no transformation is applied). If peaks incidence vary strongly across seasons, a log-tranformation is recommended \citep{WHO2017}. Irrespective of the number $m$ of historical seasons, $n = 1$ observation per season is used.
\item In the MEM method, the default choice for $f$ is the natural logarithm, while the number of observations per season is set to $n = 30/m$. (The exact specification used in the \texttt{R} package \texttt{mem} is $n = \max(1, \lfloor 30/m \rceil)$, where $\lfloor \cdot \rceil$ denotes rounding to the nearest integer.) The total number of historical observations is thus kept (approximately) fixed, ensuring a reasonably large number of reference observations
\end{itemize}
Note that the software implementation of the moving epidemic method in the \texttt{R} package \texttt{mem} is rather general and permits the user to choose $n$, $m$, and $f$ (between the identify and logarithmic function). The term \textit{moving epidemic method} could thus also be used as an umbrella term for the general procedure described above. We here use it in a more narrow sense to refer to the specific procedure described in \cite{Vega2015}, \cite{WHO2017} and reflected in the default parameters of the \texttt{mem} package.

The rationale behind the percentile-based approach is that ``about 50--60\% of the season peaks should be above the moderate threshold, $\pm 10\%$ above the high threshold and $\pm 2.5\%$ above the extraordinary threshold'' \citep{WHO2017}. Note that in most previous descriptions (e.g., \ \citealt{WHO2014, Vega2015}), the thresholds have been referred to as upper ends of one-sided confidence intervals associated with the the arithmetic (WHO method) or geometric mean (MEM) of the historical reference observations in $\mathcal{X}$. This, however, is slightly imprecise terminology as in the computations the standard deviation $\sd$ rather than the standard error $\sd/\sqrt{nm}$ of $\mean$ is used. If indeed confidence intervals were used, these would tend to get narrower the more historical seasons were available, and the thresholds for all levels would eventually converge to the arithmetic / geometric mean. This is obviously not the intended behaviour.

@@

To get a better understanding of the different settings in which the MEM and WHO methods are applied in practice we performed a literature search of articles published in English language and citing the papers \cite{Vega2015}, \cite{WHO2014} and \cite{WHO2017} until December 2020 (identified via \textit{CrossRef} and \textit{Google Scholar}). The results are summarized in Table \ref{tab:literature}. As can be seen from the numerous entries from the years 2019 and 2020, the MEM has quickly become a standard approach in the determination of intensity thresholds for influenza and other respiratory diseases. Indeed, the contributions come from numerous countries and in many cases have been co-authored by representatives of national or regional public health agencies. In most analyses, the suggested threshold levels at the 40th, 90th and 97.5th percentile are used. Variability with respect to the number $m$ of historical seasons included is considerable, with a range from three to 16 seasons. Consequently, the number $n$ of observations included per season ranges from two to ten (none of the above papers indicated a modification of the default setting $n = 30/m$ of the moving epidemic method). We only found three published applications of the WHO method, one of them providing a comparison to the thresholds from the MEM method \citep{Rguig2020}.

@@

As detailed in the previous section, the MEM and WHO methods have been applied extensively in practice and overall seem to produce thresholds which are considered useful by domain experts. A more statistical assessment, however has so far been missing. Which choice of transformation function $f$ is most reaonable is largely an empirical question. Little more can be said than that a logarithmic transformation can be helpful if the distribution of peak values is skewed. However, some statements can be made on how the value chosen for $n$ impacts the expected results for the thresholds $q_\alpha$.

We start by considering how the expectations of the empirical mean $\mean$ and standard deviation $\sd$ entering into equation \eqref{eq:def_q} depend on $n$. Denote by $\mathbf{Y}^{(j)}$ the vector of the $n$ largest transformed observations from season $j$ in decreasing order, i.e. $\mathbf{Y}^{(j)} = (Y^{(j)}_1, \dots, Y^{(j)}_n)^\top$ with $Y^{(j)}_1 \geq Y^{(j)}_2 \geq \dots \geq Y^{(j)}_n$. Now assume that $\mathbf{Y}^{(1)}, \dots, \mathbf{Y}^{(m)}$ are independent and identically distributed with mean and covariance matrix
\begin{align}
\mathbb{E}\left(\mathbf{Y}^{(j)}\right) = \boldsymbol{\mu} = \left(\begin{array}{c}
\mu_1\\
\vdots\\
\mu_n
\end{array}\right) \ \ \ \text{and} \ \ \ \text{Cov}\left(\mathbf{Y}^{(j)}\right) = \boldsymbol{\Sigma} =
\left(\begin{array}{ccc}
\sigma_{1, 1} & \cdots & \sigma_{1, n}\\
\vdots & \ddots &\vdots\\
\sigma_{n, 1} & \cdots & \sigma_{n, n}
\end{array}\right),
\end{align}
respectively, where to make notation more intuitive we also write $\sigma^2_i = \sigma_{i, i}$. It can then be shown that the expectations of $\mean$ and $\sd^2$ when using $m$ seasons and $n$ observations from each season are given by
\begin{align}
\mathbb{E}(\mean) & = \frac{1}{n} \sum_{i = 1}^n \mu_i,
\label{eq:expectation_mu}\\
\mathbb{E}(\sd^2) & = \frac{m}{mn - 1} \sum_{i = 1}^n (\sigma_{i}^2 + \mu_i^2) \ - \ \frac{1}{n(mn - 1)} \sum_{i = 1}^n \sum_{i' = 1}^n \sigma_{i,i'} \ - \ \frac{m}{n(mn - 1)}\left(\sum_{i = 1}^n \mu_i\right)^2,
\label{eq:expectation_sigma2}
\end{align}
%\begin{align}
%\mathbb{E}(\mean) & = \frac{1}{K} \sum_{i = 1}^K \mu_i,
%\label{eq:expectation_mu}\\
%\mathbb{E}(\sd^2) & = \frac{n}{nK - 1} \sum_{i = 1}^K (\sigma_{i}^2 + \mu_i^2) \ - \ \frac{1}{K(nK - 1)} \sum_{i = 1}^K \sum_{i' = 1}^K \sigma_{i,i'} \ - \ \frac{n}{K(nK - 1)}\left(\sum_{i = 1}^K \mu_i\right)^2,
%\label{eq:expectation_sigma2}
%\end{align}
respectively. The derivation is provided in Supplement \textbf{XXX}. For reasons detailed there, if the transformation function $f$ is the identity or the natural logarithm,
\begin{equation}
\mathbb{E}(q_\alpha) \approx f^{-1}\left\{\mathbb{E}(\mean) + z_\alpha \sqrt{\mathbb{E}(\sd^2)}\right\}
\label{eq:expectation_q}
\end{equation}
usually holds in good approximation in our applied setting (with $f^{-1}$ the idenity or exponential function). This implies that for $n = 1$, we have
$$
\mathbb{E}(q_\alpha) \approx f^{-1}(\mu_1 + z_\alpha \sigma_1).
$$
This is a desirable property. Loosely speaking, if in addition the $Y_{1}^{(j)}$ come from a normal distribution, the nominal exceedance probabilities for the different thresholds would on average be achieved. If a different value $n > 1$ is chosen, this will generally no longer be the case. Equations \eqref{eq:expectation_mu}--\eqref{eq:expectation_q} tell us by how much $q_\alpha$ can be expected to differ from the intended value $f^{-1}(\mu_1 + z_\alpha \sigma_1)$. By the definition of the $\mu_i$ (with $\mu_i$ the expectation of the $i$-th largest observation in a given season), $\mathbb{E}(\mean)$ decreases in $n$. In most (possibly all) settings this also translates to the $q_\alpha$. As a consequence, when choosing $n > 1$, one must expect to classify a larger number of season peaks as high or very high intensity. When choosing $n = 30/m$ as suggested for the MEM, threholds will then tend to increase the more years of data are used to compute thresholds.

@@

All four time series are displayed in Figure \ref{fig:data}. Note that we removed the pandemic 2009/2010 influenza season from each series as it is not comparable to the usual years with only seasonal influenza. We also show boxplots of the first through sixth largest observation per season (right column). In the notation introduced in Section \ref{sec:definitions}, the first boxplot thus shows observations $X_1^{(j)}, j = 1, \dots, M$, the second $X_2^{(j)}, j = 1, \dots, M$ and so on. Little surprisingly (and by definition), the average values get smaller for increasing ranks. Interestingly, they also get less dispersed, meaning that the variability between e.g.\ the sixth largest observations of the different seasons is smaller than that of the largest observations.

@@

As expected, when letting the number of observations used per season depend on the number of available seasons via $n = 30/m$, average thresholds increase in $m$. As a particularly striking example, the average threshold for high intensity for France when using a log transformation increase from %$q_{0.9} = 834, q_{0.975} = 1158$ for $m = 5, n = 6$ to $q_{0.9} = 1011, q_{0.975} = 1363$ for $m = 10, n = 3$ and $q_{0.9} = 1080, q_{0.975} = 1458$ for $m = 15, n = 2$. For $n = 1$, in which case the estimators can be interpreted as unbiased (Section \ref{sec:analytical}), we obtain averages of $q_{0.9} = 1147, q_{0.975} = 1559$.
834 for $m = 5, n = 6$ to 1011 for $m = 10, n = 3$ and 1080 for $m = 15, n = 2$. For $n = 1$, in which case the estimators can be interpreted as unbiased (Section \ref{sec:analytical}), we obtain an average of 1147. This corresponds to an increase of roughly 35\% relative to the thresholds obtained with $n = 6$. Including historical observations which are not from actual peak weeks thus leads to a considerable lowering of alarm thresholds and will necessarily increase the number of alerts for high and very high influenza activity. As an example, in Spain, applying the MEM with log transformation and $M = 5, n = 6$ leads to 25\% of all seasons exceeding the high threshold and 11\% exceeding the very high threshold. This is obviously rather far from the intended excedance probabilities of 10\% and 2.5\% for the two thresholds.

@@

The increase of thresholds as more historical data accumulates is not only a statistical tendency, but occurs with very high probability. Indeed, for all four countries, the probability that the threshold for high intensity increases when using the first ten rather than only the first five years of a sequence was around 90\%. To illutrate this aspect further, we examine the following toy example: We repeat a sequence of five seasons of ILI incidence data from France (2014--2019) four times to obtain a time series of twenty seasons. We then apply the moving epidemic method with log transformation and $n = 30/m$ using the first $m = 5, 6, \dots, 20$ years of data, thus mimicking the development of thresholds as more historical data becomes available. As can be seen from Figure \ref{fig:example_france}, this results in a quite steady increase of thresholds, despite the fact that the data in each five-year block behave exactly the same. For example, season 6 is classified as very high intensity based on seasons 1 through 5, but seasons 11 and 16 are not, even though they are identical to season 6 and the respective data used to construct thresholds follow exactly the same patterns. A similar discrepancy can be observed for season 7, which is classified as medium intensity, while seasons 12 and 17 are marked as low intensity.

@@

We provided a statistical assessment of different implementation choices in a simple and widely used framework for the computation of influenza intensity thresholds. We found that using a log transformation to intensity measures prior to the computation of thresholds empirically led to better behaved thresholds with closer to nominal exceedance rates. Concerning the question of how many observations per historical season should be included in the computation of thresholds, we provided analytical arguments and empirical evidence that the common choice $n = 30/m$ leads to too low thresholds when few historical seasons are available. We therefore recommend to adopt $n = 1$ irrespective of the number $M$ of available historical seasons. This is possible using the available implementation in the \texttt{R} package \texttt{mem} and only requires setting \texttt{i.n.max = 1} in the call of the \texttt{memmodel} to override the default value 30/m.

>>>>

\newcommand{\sd}{s}
\newcommand{\mean}{\bar{y}}

@@

% \title{A bootstrap-based comparison of ILI intensity thresholds from the moving epidemic and WHO methods}
\title{A statistical comparison of influenza intensity thresholds obtained from the moving epidemic and WHO methods}
\author{Johannes Bracher$^{1, 2}$, Jonas Littek$^1$}
\date{ \small
$^1$Karlsruhe Institute of Technology (KIT), Chair of Statistics and Econometrics\\
$^2$Heidelberg Institute for Theoretical Studies, Computational Statistics Group}

@@

\abstract{The moving epidemic method (MEM) and the WHO method are widely used approaches to determine intensity levels for seasonal influenza and influenza-like illness (ILI). They are conceptually similar, but differ in two aspects. Firstly, the MEM involves a log-transformation of incidence data, while the WHO method typically operates on the original scale of the data. Secondly, the MEM method by default uses more than one observation from each past season, with a fixed total number of observations to include. The WHO method, on the other hand, uses only the single highest value from each season to determine peak intensity thresholds. We provide statistical arguments and a simulation study to assess the impact of these choices on the resulting intensity thresholds and empirical exceedance proportions. The simulation study is based on a re-sampling approach using historical ILI data from France, Spain, Switzerland and the United States. We find that using data on the original scale leads to a rather large proportion of season peaks being classified as high or very high intensity, which can be mitigated to some degree by using a logarithmic transformation. When fixing the total number of observations to be included as in the MEM, intensity thresholds tend to increase as more historical data becomes available. When data are only available for few seasons, there is a high chance of classifying a new season as high or very high intenstity. We therefore advocate always using one observation per season with a log transformation, i.e.\ a hybrid between the default settings of the MEM and WHO methods. This adaptation is possible within an existing software implementation of the MEM approach and thus easy to apply.}

@@

Following the 2009 influenza H1N1 pandemic, the need for a rapid assessment tool for influenza intensity was recognized. In a report, the \textit{Review Committee on the Functioning of the
International Health Regulations and on Pandemic Influenza (H1N1)} recommended that member states apply and evaluate their methods for severity assessment every year, thus improving their pandemic preparedness \citep[p.118]{WHO2011}. In the subsequent \textit{WHO Pandemic Infuenza Severity Assessment (PISA)} guideline \citep{WHO2017}, two specific statistical methods are recommended to determine influenza intensity thresholds. Firstly, the so-called \textit{WHO method}, already described in a previous document \citep{WHO2014} and in more detail in the PISA guide; and secondly, the \textit{moving epidemic method} (MEM; \citealt{Vega2013, Vega2015}) which is also recommended by the European Centre for Disease Prevention and Control (e.g., \citealt{ECDC2017}).

In particular the MEM approach has in the meantime been adopted by numerous national public health agencies (see references in Section \ref{sec:recent_applications}) and is considered to deliver practically useful thresholds. Its application has been facilitated by the availablility of a user-friendly and well-documented \texttt{R} package (\texttt{mem}; \citealt{Lozano2020}), which can even be used via a graphical user interface \citep{Lozano2018}. The statistical properties of the MEM and WHO methods, however, have not yet been studied in detail. We obtain some analytical results, complemented by simulation experiments. In order to closely mimic real influenza patterns, the latter are based on bootstrap re-sampling of historical surveillance data. In this way, we provide an improved understanding and intuition for the consequences of different implementation choices.

The article is structured as follows. Section \ref{sec:definitions} provides concise definitions of the MEM and WHO methods, highlighting two differences between these conceptually similar approaches. Section \ref{sec:recent_applications} consists of an overview of published applications of the MEM and WHO methods. This is followed by a brief examination of statistical properties of the two methods in Section \ref{sec:analytical}. In Section \ref{sec:simulation} we describe our simulation approach and the results based on historical data from France, Spain, Switzerland and the United States. Section \ref{sec:discussion} concludes with a discussion.

@@

In the following we focus on the construction of thresholds  for peak intensity, framing the MEM and WHO methods as different special cases of the same general approach. Details on the construction of baseline epidemic thresholds from pre-season observations are omitted. We assume that thresholds are based on influenza intensity data (typically weekly incidences) from $m$ past seasons and applied to the ($m + 1$)-th season. \cite{Vega2015} recommend to use $5 \leq m \leq 10$ seasons for the MEM to ensure a recent data basis. Thresholds for a given intensity measure $X$ are then obtained via the following steps. In the following we will call $X$ the incidence for simplicity, even if it can also be a different measure.

\begin{enumerate}
\item Within each historical season $j = 1, \dots, m$ order all observations in decreasing order, denoting the $i$-th largest observation from season $j$ by $x^{(j)}_i$.
\item Select the $n$ largest observations from each of the $m$ past seasons to construct a reference set $\mathcal{X} = \{x_i^{(j)}: j = 1, \dots, m; i = 1, \dots, n\}$.
\item Apply an (invertible) transformation $y_i^{(j)} = f(x_i^{(j)})$ to all members of the reference set $\mathcal{X}$ to obtain a reference set $\mathcal{Y}$ of transformed historical observations.
\item Assume that the transformed values in $\mathcal{Y}$ come from a normal distribution and compute estimates $\mean, \sd$ of its mean and standard deviation.
\item Define intensity thresholds on the transformed scale as quantiles of the normal distribution N$(\mean, \sd^2)$, i.e.\ compute
\begin{equation}
q_{Y, \alpha} = \mean + z_\alpha \sd, \label{eq:def_q}
\end{equation}
where $z_\alpha$ is the $\alpha$ quantile of the standard normal distribution. A common choice for both methods \citep{WHO2017}, also implemented as the default in the \texttt{R} package \texttt{mem} \citep{Lozano2020} is to use
\begin{itemize}
\item the 40th percentile $q_{Y, 0.4} = \mean - 0.25 \sd$ as the threshold between low and medium intensity;
\item the 90th percentile $q_{Y, 0.9} = \mean + 1.28 \sd$ as the threshold between medium and high intensity;
\item the 97.5th percentile $q_{Y, 0.975} = \mean + 1.96\sd$ as the threshold between high and very high intensity.
\end{itemize}
\item Obtain thresholds on the original scale by applying the inverse transformation, i.e. setting $q_{0.4} = f^{-1}(q_{Y, 0.4})$, $q_{0.9} = f^{-1}(q_{Y, 0.9})$, $q_{0.975} = f^{-1}(q_{Y, 0.975})$.
\end{enumerate}

\noindent The MEM and WHO methods are special cases of this procedure with the following specifications:
\begin{itemize}
\item In the WHO method, $n = 1$ observation per season is used irrespective of the number $m$ of historical seasons. The standard procedure is to apply no transformation (i.e.\ $f$ is set to the identity function). If peak incidences vary strongly across seasons, a log-tranformation is recommended \citep{WHO2017}. 
\item In the MEM method, the default choice for $f$ is the natural logarithm, while the number of observations per season is set to $n = 30/m$. (The exact specification used in the \texttt{R} package \texttt{mem} is $n = \max(1, \lfloor 30/m \rceil)$, where $\lfloor \cdot \rceil$ denotes rounding to the nearest integer.) The total number of historical observations is thus kept approximately fixed, ensuring a reasonably large number of reference observations.
\end{itemize}
The software implementation of the moving epidemic method in the \texttt{R} package \texttt{mem} is rather general and permits the user to choose $n$, $m$, and $f$ (the identify function coresponding to \texttt{i.type.intensity	= 5} and logarithmic function to \texttt{i.type.intensity = 6}). The term \textit{moving epidemic method} could thus also be used as an umbrella term for the general procedure described above. We use it here in a more narrow sense to refer to the specific procedure described in \cite{Vega2015}, \cite{WHO2017} and reflected in the default parameters of the \texttt{mem} package.

The rationale behind the percentile-based approach is that ``about 50--60\% of the season peaks should be above the moderate threshold, $\pm 10\%$ above the high threshold and $\pm 2.5\%$ above the extraordinary threshold'' \citep{WHO2017}. Note that in most previous descriptions (e.g., \ \citealt{WHO2014, Vega2015}), the thresholds have been referred to as upper ends of one-sided confidence intervals associated with the the arithmetic (WHO method) or geometric mean (MEM) of the historical reference observations in $\mathcal{X}$. This, however, is imprecise terminology as in the computations the standard deviation $\sd$ is used rather than the estimated standard error $\sd/\sqrt{nm}$ (see the default settings in the R package \texttt{mem} and the description of the WHO method in \citealt{WHO2014}, p. 69). We note that the use of actual confidence intervals is in fact also possible using the \texttt{mem} package (\texttt{i.type.intensity = 1} for the geometric mean, \texttt{i.type.intensity = 2} for the arithmetic mean). However, under this choice, the thresholds for all levels eventually converge to the arithmetic or geometric mean if sufficient historical data are available (as $\sd/\sqrt{nm}$ gets smaller and smaller). This is obviously not the intended behaviour. To illustrate this aspect we also provide simulation studies on these two choices, but these have been deferred to Appendix \textbf{XXX}.

@@

To get a better understanding of the different settings in which the MEM and WHO methods are applied in practice we performed a literature search of articles published in English and citing the papers \cite{Vega2015}, \cite{WHO2014} and \cite{WHO2017} until December 2020 (identified via \textit{CrossRef} and \textit{Google Scholar}). The results are summarized in Table \ref{tab:literature}. As can be seen from the numerous entries from the years 2019 and 2020, the MEM has quickly become a standard approach in the determination of intensity thresholds for influenza and other respiratory diseases. Indeed, the contributions come from numerous countries and in many cases have been co-authored by representatives of national or regional public health agencies. In most analyses, the suggested threshold levels at the 40th, 90th and 97.5th percentile as described in Section \ref{sec:definitions} are used. Variability with respect to the number $m$ of historical seasons included is considerable, with a range from three to 16 seasons. Consequently, the number $n$ of observations included per season ranges from two to ten (none of the above papers indicated a modification of the default setting $n = 30/m$ of the moving epidemic method). We only found three published applications of the WHO method, one of them providing a comparison to the thresholds from the MEM method \citep{Rguig2020}.

@@

As detailed in the previous section, the MEM and WHO methods have been applied extensively in practice and overall seem to produce thresholds that are considered useful by domain experts. A more statistical assessment, however has so far been missing. Determining which choice of transformation function $f$ is most reaonable is largely an empirical question. Little more can be said than that a logarithmic transformation can be helpful if the distribution of peak values is skewed. However, some statements can be made on how the value chosen for $n$ impacts the expected results for the thresholds $q_\alpha$.

We start by considering how the expectations of the empirical mean $\bar{Y}$ and standard deviation $S$ entering into equation \eqref{eq:def_q} depend on $n$ (note that we now capitalize these as we consider the random variables rather than concrete observed values). Denote by $\mathbf{Y}^{(j)}$ the vector of the $n$ largest transformed incidences from season $j$ in decreasing order, i.e. $\mathbf{Y}^{(j)} = (Y^{(j)}_1, \dots, Y^{(j)}_n)^\top$ with $Y^{(j)}_1 \geq Y^{(j)}_2 \geq \dots \geq Y^{(j)}_n$. Now assume that $\mathbf{Y}^{(1)}, \dots, \mathbf{Y}^{(m)}$ are independent and identically distributed with mean and covariance matrix
\begin{align}
\mathbb{E}\left(\mathbf{Y}^{(j)}\right) = \boldsymbol{\mu} = \left(\begin{array}{c}
\mu_1\\
\vdots\\
\mu_n
\end{array}\right) \ \ \ \text{and} \ \ \ \text{Cov}\left(\mathbf{Y}^{(j)}\right) = \boldsymbol{\Sigma} =
\left(\begin{array}{ccc}
\sigma_{1, 1} & \cdots & \sigma_{1, n}\\
\vdots & \ddots &\vdots\\
\sigma_{n, 1} & \cdots & \sigma_{n, n}
\end{array}\right),
\end{align}
respectively, where to make notation more intuitive we also write $\sigma^2_i = \sigma_{i, i}$. It can then be shown that the expectations of $\mean$ and $\sd^2$ when using $m$ seasons and $n$ observations from each season are given by
\begin{align}
\mathbb{E}(\mean) & = \frac{1}{n} \sum_{i = 1}^n \mu_i,
\label{eq:expectation_mu}\\
\mathbb{E}(\sd^2) & = \frac{m}{mn - 1} \sum_{i = 1}^n (\sigma_{i}^2 + \mu_i^2) \ - \ \frac{1}{n(mn - 1)} \sum_{i = 1}^n \sum_{i' = 1}^n \sigma_{i,i'} \ - \ \frac{m}{n(mn - 1)}\left(\sum_{i = 1}^n \mu_i\right)^2,
\label{eq:expectation_sigma2}
\end{align}
%\begin{align}
%\mathbb{E}(\mean) & = \frac{1}{K} \sum_{i = 1}^K \mu_i,
%\label{eq:expectation_mu}\\
%\mathbb{E}(\sd^2) & = \frac{n}{nK - 1} \sum_{i = 1}^K (\sigma_{i}^2 + \mu_i^2) \ - \ \frac{1}{K(nK - 1)} \sum_{i = 1}^K \sum_{i' = 1}^K \sigma_{i,i'} \ - \ \frac{n}{K(nK - 1)}\left(\sum_{i = 1}^K \mu_i\right)^2,
%\label{eq:expectation_sigma2}
%\end{align}
respectively. The derivation is provided in Supplement \textbf{XXX}. For reasons detailed there, if the transformation function $f$ is the identity or the natural logarithm,
\begin{equation}
\mathbb{E}(q_\alpha) \approx f^{-1}\left\{\mathbb{E}(\mean) + z_\alpha \sqrt{\mathbb{E}(\sd^2)}\right\}
\label{eq:expectation_q}
\end{equation}
usually holds in good approximation in our applied setting (with $f^{-1}$ the idenity or exponential function). This implies that for $n = 1$, we have
$$
\mathbb{E}(q_\alpha) \approx f^{-1}(\mu_1 + z_\alpha \sigma_1).
$$
This is a desirable property. Loosely speaking, if in addition the $Y_{1}^{(j)}$ come from a normal distribution, the nominal exceedance probabilities for the different thresholds would be achieved on average. If a different value $n > 1$ is chosen, this will generally no longer be the case. Equations \eqref{eq:expectation_mu}--\eqref{eq:expectation_q} tell us by how much $q_\alpha$ can be expected to differ from the intended value $f^{-1}(\mu_1 + z_\alpha \sigma_1)$. By the definition of the $\mu_i$ (with $\mu_i$ the expectation of the $i$-th largest observation in a given season), $\mathbb{E}(\mean)$ decreases in $n$. In most (possibly all) settings this also translates to the $q_\alpha$. As a consequence, when choosing $n > 1$, one must expect to classify a larger number of season peaks as high or very high intensity. When choosing $n = 30/m$ as suggested for the MEM, threholds will then tend to increase the more years of data are used to compute thresholds.

@@

All four time series are displayed in Figure \ref{fig:data}. Note that we removed the pandemic 2009/2010 influenza season from each series as it is not comparable to the usual years with only seasonal influenza. We also show boxplots of the first through sixth largest observation per season (right column). In the notation introduced in Section \ref{sec:definitions}, the first boxplot thus shows observations $X_1^{(j)}, j = 1, \dots, M$, the second $X_2^{(j)}, j = 1, \dots, M$ and so on. Not surprisingly (and in fact by definition), the average values get smaller for increasing ranks. Interestingly, they also get less dispersed, meaning that the variability among e.g.\ the sixth largest observations of the different seasons is smaller than that among the largest observations.

@@

As expected, when letting the number of observations used per season depend on the number of available seasons via $n = 30/m$, average thresholds increase in $m$. As a particularly striking example, the average threshold for high intensity for France when using a log transformation increase from %$q_{0.9} = 834, q_{0.975} = 1158$ for $m = 5, n = 6$ to $q_{0.9} = 1011, q_{0.975} = 1363$ for $m = 10, n = 3$ and $q_{0.9} = 1080, q_{0.975} = 1458$ for $m = 15, n = 2$. For $n = 1$, in which case the estimators can be interpreted as unbiased (Section \ref{sec:analytical}), we obtain averages of $q_{0.9} = 1147, q_{0.975} = 1559$.
834 for $m = 5, n = 6$ to 1011 for $m = 10, n = 3$ and 1080 for $m = 15, n = 2$. For $n = 1$, in which case the estimators can be interpreted as unbiased (Section \ref{sec:analytical}), we obtain an average of 1147. This corresponds to an increase of roughly 35\% relative to the thresholds obtained with $n = 6$. Including historical observations which are not from actual peak weeks thus leads to a considerable lowering of alarm thresholds and will necessarily increase the number of alerts for high and very high influenza activity. As an example, in Spain, applying the MEM with log transformation and $M = 5, n = 6$ leads to 25\% of all seasons exceeding the high threshold and 11\% exceeding the very high threshold. This is obviously rather far from the intended excedance probabilities of 10\% and 2.5\% for the two thresholds.\todo{Add France low/medium}

@@

The increase of thresholds as more historical data accumulate is not only a statistical tendency, but occurs with very high probability. Indeed, for all four countries, the probability that the threshold for high intensity increases when using the first ten rather than only the first five years of a sequence was around 90\%. To illutrate this aspect further, we examine the following toy example: We repeat a sequence of five seasons of ILI incidence data from France (2014--2019) four times to obtain a time series of twenty seasons. We then apply the moving epidemic method with log transformation and $n = 30/m$ using the first $m = 5, 6, \dots, 20$ years of data, thus mimicking the development of thresholds as more historical data become available. As can be seen from Figure \ref{fig:example_france}, this results in a quite steady increase of thresholds, despite the fact that the data in each five-year block behave exactly the same. For example, season 6 is classified as very high intensity based on seasons 1 through 5, but seasons 11 and 16 are not, even though they are identical to season 6 and the respective data used to construct thresholds follow exactly the same patterns. A similar discrepancy can be observed for season 7, which is classified as medium intensity, while seasons 12 and 17 are marked as low intensity.

@@

We provided a statistical assessment of different implementation choices in a simple and widely used framework for the computation of influenza intensity thresholds. We found that log-transforming intensity measures prior to the computation of thresholds empirically led to better behaved thresholds with closer to nominal exceedance rates. Concerning the question of how many observations per historical season should be included in the computation of thresholds, we provided analytical arguments and empirical evidence that the common choice $n = 30/m$ leads to too low thresholds when few historical seasons are available. We therefore recommend adopting $n = 1$ irrespective of the number $m$ of available historical seasons. This is possible using the available implementation in the \texttt{R} package \texttt{mem} and only requires setting \texttt{i.n.max = 1} in the call of the \texttt{memmodel} to override the default value 30/m.
